Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-16 draft.tex
.

@@

\subsubsection{Low-Density Neighbourhoods correlate with strong outgoing Connections}
In Section \ref{Section_Mean_outgoing_Weights}, we stated that diffusive homeostasis allowed the network do develop few neurons with exceptionally strong outgoing weights. Furthermore, we found this effect to be present in neurons with an highly above-average firing rate. Moreover, the outcomes presented in Section \ref{Section_Rand_Mat_vs_Sim} suggest that the steady state of firing rates is strongly influenced by the spatial structure of excitatory neurons. Combining these two qualities led us the conclusion that one should be able to observe some relation between spatial structure and the emergence of "driver neurons". More specific, regions of neurons with above average firing rates should develop stronger outgoing connections. To get an intuitive understanding of what characterizes regions of above-average firing rates, we interpreted the homeostatic condition in the high-neuron-density limit as an approximation of a continuous integral equation:
\begin{equation}
\int d^2 x' \rho(\mathbf{x}') r(\mathbf{x}') \psi(|\mathbf{x}-\mathbf{x}'|) = NO_0 \; , \; \forall \, \mathbf{x}
\label{Density_Limit_Hom}
\end{equation}
where $\mathrm{\rho(\mathbf{x})}$ is the local neuron density and $\mathrm{r(\mathbf{x})}$ the local firing rate. Here, we assumed that our tissue is of infinite size and thus the integral is carried out over $\mathbb{R}^2$. In this form, the solution for $\mathrm{r(\mathbf{x})}$ is quite trivial:
\begin{equation}
r(\mathbf{x}) = \frac{NO_0}{\rho(\mathbf{x}) \int d^2 x' \psi(|\mathbf{x}'|)} \propto \frac{1}{\rho(\mathbf{x})} \; .
\label{Density_Limit_Hom_Solution}
\end{equation}
Combined with the fact that the logarithm of outgoing excitatory weights correlates with excitatory firing rate (see Figure \ref{Out_Weight_vs_F}), this led us to the conclusion that low density regions should contain neurons with strong outgoing weights and vice versa. To test this hypothesis, we approximated the local neuron density by means of a a gaussian kernel with $\mathrm{\sigma = 50\, \mu m}$. Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} shows the results. Despite a certain amount of randomness, our prediction was verified. 
\begin{figure}
\includegraphics[width=\textwidth]{../../plots/scatter_density_out_weights.png}
\includegraphics[width=\textwidth]{../../plots/scatter_density_out_weights_points.png}
\caption{(A) and (B): Scatter-plot of inverse of local neuron density (calculated by convolution with a gaussian kernel with $\mathrm{\sigma = 50\, \mu m}$) and decadic logarithm of outgoing excitatory weights. Note the spatial correlation in high/low-density regions. (C) and (D): Decadic logarithm of outgoing weights correlates with inverse neuron density (corr. coeff. of 0.513).}
\label{Inverse_Dens_vs_Sum_Out_Weights}
\end{figure}

@@

In general, diffusive homeostasis is regarded as a possibility of maintaining an overall level of activity, while allowing single neurons greater variability compared to single-neuron homeostasis, be it a single or individual target rates. According to Sweeney et al., this allows for the encoding of heterogeneous input by differences in the resulting excitatory activity. For this effect to be persistent, one should be able to alter the homeostatic fixed point by changes in external input.
Our results did not support this. Rather, by analysing network activity and comparing it to our analytic model, we found evidence that the fixed point of network activity is strongly determined and can be well predicted by the spatial configuration of excitatory neurons. However, the validity of this prediction declined for larger diffusion constants and completely vanished for instant diffusion. In a strict sense, only instant diffusion allows for an infinite number of activity configurations that fulfil the homeostatic condition. However, if diffusion in a real biological system happens at a finite but fast speed compared to the adaptation of intrinsic excitability, one could arguably expect it to be equivalent to instant diffusion. In our simulation, we explained the fact that we saw a steady decline of correlation in Figure \ref{Corr_Coeff_vs_D} by the decreasing capability of the homeostatic mechanism to quickly enough react to structural synaptic changes that alter the recurrent excitatory inputs and, in consequence, excitatory firing rates.

As an indirect consequence of spatial structure determining activity, we found that neurons with a low-density neighbourhood have a higher chance of developing strong outgoing weights. Of course, this raises the question, whether one can expect such an effect to be observable in the nervous system.

>>>>

As an additional test, we applied external poisson input for $\mathrm{t\geq 300\, s}$ to two excitatory subgroups to see whether changes in the firing rate of these subgroups is levelled out by homeostasis or retained. Members of these subgroups were picked randomly and independent of their position.
\begin{figure}
\begin{center}
%% plot_external_input_switch_activity.py
\includegraphics[width=0.7\textwidth]{../../plots/external_input_switch.png}
\end{center}
\caption{Excitatory population firing rate of subgroups subject to external poisson input for $\mathrm{t\geq 300\, s}$. Red and green subgroups consist of 100 randomly picked exc. neurons each. The remaining 300 neurons (blue) received no external input.}
\label{Ext_Input_Switch}
\end{figure}

@@

\subsubsection{Low-Density Neighbourhoods correlate with strong outgoing Connections}
In Section \ref{Section_Mean_outgoing_Weights}, we stated that diffusive homeostasis allowed the network do develop few neurons with exceptionally strong outgoing weights. Furthermore, we found this effect to be present in neurons with an highly above-average firing rate. Moreover, the outcomes presented in Section \ref{Section_Rand_Mat_vs_Sim} suggest that the steady state of firing rates is strongly influenced by the spatial structure of excitatory neurons. Combining these two qualities led us the conclusion that one should be able to observe some relation between spatial structure and the emergence of "driver neurons". More specific, regions of neurons with above average firing rates should develop stronger outgoing connections. To get an intuitive understanding of what characterizes regions of above-average firing rates, we interpreted the homeostatic condition in the high-neuron-density limit as an approximation of a continuous integral equation:
\begin{equation}
\int d^2 x' \rho(\mathbf{x}') r(\mathbf{x}') \psi(|\mathbf{x}-\mathbf{x}'|) = NO_0 \; , \; \forall \, \mathbf{x}
\label{Density_Limit_Hom}
\end{equation}
where $\mathrm{\rho(\mathbf{x})}$ is the local neuron density and $\mathrm{r(\mathbf{x})}$ the local firing rate. Here, we assumed that our tissue is of infinite size and thus the integral is carried out over $\mathbb{R}^2$. In this form, the solution for $\mathrm{r(\mathbf{x})}$ is quite trivial:
\begin{equation}
r(\mathbf{x}) = \frac{NO_0}{\rho(\mathbf{x}) \int d^2 x' \psi(|\mathbf{x}'|)} \propto \frac{1}{\rho(\mathbf{x})} \; .
\label{Density_Limit_Hom_Solution}
\end{equation}
Combined with the fact that the logarithm of outgoing excitatory weights correlates with excitatory firing rate (see Figure \ref{Out_Weight_vs_F}), this led us to the conclusion that low density regions should contain neurons with strong outgoing weights and vice versa. To test this hypothesis, we approximated the local neuron density by means of a a gaussian kernel with $\mathrm{\sigma = 50\, \mu m}$. Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} shows the results. Despite a certain amount of randomness, our prediction was verified. 
\begin{figure}
%% density_vs_sum_out_weights.py
\includegraphics[width=\textwidth]{../../plots/scatter_density_out_weights.png}
%% density_vs_sum_out_weights_points.py
\includegraphics[width=\textwidth]{../../plots/scatter_density_out_weights_points.png}
\caption{(A) and (B): Scatter-plot of inverse of local neuron density (calculated by convolution with a gaussian kernel with $\mathrm{\sigma = 50\, \mu m}$) and decadic logarithm of outgoing excitatory weights. Note the spatial correlation in high/low-density regions. (C) and (D): Decadic logarithm of outgoing weights correlates with inverse neuron density (corr. coeff. of 0.513).}
\label{Inverse_Dens_vs_Sum_Out_Weights}
\end{figure}

@@

In general, diffusive homeostasis is regarded as a possibility of maintaining an overall level of activity, while allowing single neurons greater variability compared to single-neuron homeostasis, be it a single or individual target rates. According to Sweeney et al., this allows for the encoding of heterogeneous input by differences in the resulting excitatory activity. For this effect to be persistent, one should be able to alter the homeostatic fixed point by changes in external input.
Our results did not support this. As shown in Figure \ref{Ext_Input_Switch}, we found that diffusive homeostasis does not maintain separability of external input by the resulting firing rates. Rather, by analysing network activity and comparing it to our analytic model, we found evidence that the fixed point of network activity is strongly determined and can be well predicted by the spatial configuration of excitatory neurons. However, the validity of this prediction declined for larger diffusion constants and completely vanished for instant diffusion. In a strict sense, only instant diffusion allows for an infinite number of activity configurations that fulfil the homeostatic condition. Presumably however, if diffusion in a real biological system happens at a finite but fast speed compared to the adaptation of intrinsic excitability, one could arguably expect it to be equivalent to instant diffusion. In our simulation, we explained the fact that we saw a steady decline of correlation in Figure \ref{Corr_Coeff_vs_D} by the decreasing capability of the homeostatic mechanism to quickly enough react to structural synaptic changes that alter the recurrent excitatory inputs and, in consequence, excitatory firing rates.

As an indirect consequence of spatial structure determining activity, we found that neurons with a low-density neighbourhood have a higher chance of developing strong outgoing weights. Of course, this raises the question, whether one can expect such an effect to be observable in the nervous system. To our knowledge, no experimental study exists that directly relates fluctuations of neuronal densities to the existence of highly influential excitatory neurons. Ivenshitz and Segal studied the influence of neuronal density onto neuronal activity in a culture of hippocampal neurons \cite{Ivenshitz_2010}. Lower densities resulted in a more irregular activity, with stronger but less frequent bursting events. However, since our network was tuned to exhibit non-bursting activity, it is hard to draw any conclusions from this result.

A major problem of the implemented model is the usage of quasi-point sources of nitric oxide, since it leads to a clear single homeostatic fixed point. Future modifications of the homeostatic model might integrate a wider-spread kernel of nNOS, allowing for a larger amount of states of activity fulfilling the homeostatic target. In fact, experimental and theoretical studies have suggested that NO is synthesized across wider spread cell structures, leading to a more even distribution of NO \cite{Philippides_2005}.

While the implemented mathematical description of diffusion (Equation \eqref{NO_dyn}) is physically based, one can question the validity of the implementation of threshold adaption (Equation \eqref{Theta_dyn}): In principle, it does not impose any bounds onto the threshold value, which, from a biological point of view, seems implausible.
